A Comparison of Regression models for Cylindrical data in Psychology.

### 2018-07-02 Manuscript/ManuscriptMarkdownDifferentSE.Rmd
Manuscript Update

---
title: "Working title: Regression models for Cylindrical data."
author: |
  | Jolien Cremers^[Corresponding author: j.cremers@uu.nl] $^1$, Helena J.M. Pennings$^{2,3}$,  Christophe Ley $^{4}$
  | $^1$Department of Methodology and Statistics, Utrecht University
  | $^2$TNO
  | $^3$Department of Education, Utrecht University
  | $^4$Department of Applied Mathematics, Computer Science and Statistics, Ghent University

bibliography: CircularData.bib
csl: apa.csl
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
  html_document: default
  word_document: default
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{multirow}
  - \usepackage{appendix}
  - \usepackage{color}
  - \usepackage{hyperref}
  - \usepackage{subcaption}
keywords: bla vla
documentclass: article
pandoc_args: --natbib
abstract: |
  This is the abstract
---
```{r inlude = FALSE, echo = FALSE}

@@

```{r, echo = FALSE}
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
```

@@

Cylindrical data are data that consist of a linear component and a directional 
component (a set of angles). In this paper, the directional component is 
circular meaning that it consists of a single angle instead of a set of angles. 
Figure \ref{} \textcolor{red}{NOG MAKEN} shows the difference between a circular variable (left) and a 
linear variable (right). The most obviuous difference is that on a circular 
scale the datapoints 0$^\circ$ and 360$^\circ$ are connected and in fact 
represent the same number while on a linear scale the two ends, $-\infty$ and 
$\infty$ are not connected. Cylindrical data occur in several fields of 
research, such as for instance in meteorology [@garcia2013exploring], ecology 
[@garcia2014test] or marine research [@lagona2015hidden]. Several types of data 
in psychological research are also of a cylindrical nature. For example, in 
research on human navigation in the field of cognitive psychology both distance,
a linear variable, and direction, a circular variable, of movement are of 
interest [@chrastil2017rotational]. In eye-tracking research, which is used for 
investigating cognitive processes we can also speak of cylindrical data. When 
measuring eye-movements we speak of the eye-movements themselves, the saccades, 
and fixations, the periods of time between movements when the eyes are still and
looking at one point. Of the saccades both the direction, a circular variable, 
and the duration, a linear variable, are of interest (for a review of 
eye-tracking research see @rayner200935th). Data from circumplex measurement
instruments, e.g. the interpersonal circumplex as used in personality
psychology, are also of a cylindrical nature (see Section \ref{Example} for a
more detailed explanation). In this paper we will discuss how a correct
statistical treatment of such cylindrical data can lead to new insights. In
particular we will show how cylindrical models pave the way for circular-linear
and linear-circular regression. We will do this for a motivating example from
the field of educational psychology. We will refer to this example data as the
teacher data. In this example, apart from modelling the relation between the
linear and circular component of a cylindrical variable we would also like to
predict the two components from a set of covariates in a regression model. The
teacher data will be further introduced in Sections \ref{Example} and
\ref{DataAnalysis}.

As is the case for circular data, the analysis of cylindrical data requires 
special methods. For an overview of methods for circular data see
@mardia2000directional and @jammalamadaka2001topics and for modern treatments
@ley2017modern. In the literature several methods have been put forward to
model the relation between the linear and circular component of a cylindrical 
variable. Some of these are based on regressing the linear component onto the 
circular component using the following type of relation: $$y = \beta_0 + 
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,$$ where $y$ is the linear
component and $\theta$ the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a 
different way, e.g. by specifying a multivariate model for several linear and 
circular variables and modelling their covariance matrix 
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. 
For example, @abe2017tractable introduce a cylindrical distribution  based on a 
Weibull distribution for the linear component and a sine-skewed von Mises 
distribution for the circular component and link these through their shape and 
concentration parameters respectively. However, none of the methods that have 
been proposed thus far include additional covariates on which both the circular 
and linear component are regressed.

Our aim in this paper is to fill this gap in the literature by  adapting four of
the existing methods for the analysis of cylindrical data in such a way that
they include a regression of both the linear and circular component of a
cylindrical variable onto a set of covariates. First however, we will introduce
the teacher data in Section \ref{Example}. In Section \ref{Models} we introduce 
the four modified models for cylindrical data that we use to analyze the data from
the motivating example. We also choose a model selection criterion to compare
the models. The teacher data will be analysed in Section \ref{DataAnalysis}. The
paper will be concluded with a discussion in Section \ref{Discussion}. The
Appendix contains technical details.

@@

The motivating example for this article comes from the field of educational 
psychology and was collected for the studies of @pennings2017interpersonal,
@Claessens2016side and @vanderWant2015role. These studies concern research on 
classroom climate. An indicator of classroom climate is the students'
perceptions of their teachers' interpersonal behavior. These interpersonal
perceptions, both in educational psychology as well as in other areas of
psychology, can be measured using circumplex measurement instruments (see
@horowitz2010handbook for an overview of many such instruments).

The Questionnaire on Teacher Interaction (QTI) [@wubbels2006interpersonal] is
one such circumplex measurement instrument that can be used to study student
perceptions of their teachers' interpersonal behavior. This instrument contains
items that load on two dimensions: Agency and Communion. Agency refers to the
degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two 
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. Different parts of
this space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. We call the two-dimensional space the interpersonal circle (IPC).
The idea is that the IPC is ``a continuous order with no beginning or end''
[@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering and
the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular variable.

\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{IPC-T.png}
\caption{The interpersonal circle for teachers (IPC-T). The words presented in
the circumference of the circle are anchor words to describe the type of
behavior located in each part of the IPC.}
\label{QTI}
\end{figure}

@Cremers2018Assessing show how data from the IPC can be
considered circular data and analyzed as such using a regression model. The 
two-dimension scores Agency and Communion can be converted to a circular score 
using the two-argument arctangent function in Equation \eqref{PredVal}, where 
$A$ represents a score on the Agency dimension and $C$ represents a score on the
Communion dimension. The resulting circular variable $\theta$ can then be
modeled and for example be regressed onto covariates in a circular regression 
model. However, when two-dimensional data are converted to the circle we lose 
some information, namely the length of the two-dimensional vector $(A, C)$, 
\emph{i.e.}, its euclidean norm $\mid\mid (A, C) \mid\mid$. This length 
represents the strength of the type of interpersonal behavior a teacher shows 
towards his/her students. In a cylindrical model we are able to incorporate this
information, and model a circular variable $\theta$ together with a linear 
variable corresponding to $\mid\mid (A, C) \mid\mid$. This will lead to an 
improved analysis of data from the IPC.  In the next section we will introduce 
several models that we can use for a more accurate and more informative regression
analysis on the teacher data. More information on the teacher data will be given
in Section \ref{DataAnalysis}.

@@

#center and remove missings for self-efficacy
Dat <- Dat %>% drop_na(Efficacy_SE)
Dat$SEc <- Dat$Efficacy_SE- mean(Dat$Efficacy_SE)

@@

In this section we will introduce four cylindrical regression models. We will 
extend the models from @mastrantonio2018joint and @abe2017tractable to contain 
predictors for the linear and circular outcomes in the cylindrical model. 
Additionally we introduce two models where both variates $\Theta$ and $Y$ are
predicted by covariates and that model the relation between the linear and
circular outcome as follows (following @wang2015joint and
@mastrantonio2015bayesian):

@@

In both of these models the relation between the circular variable $\Theta \in
[0, 2\pi)$ and linear variable $Y \in (-\infty, + \infty)$ is specified as in
Equation \ref{circlinlink}. The distribution of $Y$ conditional on $\Theta$ is
then given as:

@@

In a regression setup the outcome $\theta_i$ for each individual $i = 1,
\dots, n$, where $n$ is the sample size, is generated independently from
Equation (\ref{pnreg}). The mean vector $\boldsymbol{\mu}_i \in \mathbb{R}^2$ is 
defined as $\boldsymbol{\mu}_i = \boldsymbol{z}_i\boldsymbol{B}$. The vector 
$\boldsymbol{z}_i$ is a a vector of $p$ covariate values and $\boldsymbol{B} = 
(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})$ contain the regression 
coefficients. Note however that the dimensions of $\boldsymbol{\beta}^{I}$ and 
$\boldsymbol{\beta }^{I}$ need not necessarily be the same and we are thus 
allowed to have a different set of predictor variables and vectors 
$\boldsymbol{z}_i^I$ and $\boldsymbol{z}_i^{II}$ for the two components of 
$\boldsymbol{\mu}_i$.

@@

where $\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2  & \rho\\ \rho & 1 
\end{bmatrix}$, $\boldsymbol{u}= \begin{bmatrix} \cos \theta \\ \sin \theta 
\end{bmatrix}$. In a regression setup the outcome $\theta_i$ for each individual
$i = 1, \dots, n$, where $n$ is the sample size, is generated independently
from Equation (\ref{gpnreg}). The mean vector $\boldsymbol{\mu}_i \in 
\mathbb{R}^2$ is defined as $\boldsymbol{\mu}_i = 
\boldsymbol{z}_i(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})$. The vector 
$\boldsymbol{z}_i$ is a a vector of $p$ covariate values and each 
$\boldsymbol{\beta}^{k}$ is a vector with two regression coefficients, one for 
each of the two components of $\boldsymbol{\mu}_i$. Note that for the CL-GPN 
model we do need to have the same predictors for both components of 
$\boldsymbol{\mu}_i$.

Both cylindrical models introduced in this section are estimated using MCMC 
methods. These methods are based on @nunez2011bayesian, @wang2012directional and
@hernandez2016general for the regression of the circular outcome. A detailed
description of the Bayesian estimation and MCMC samplers can be found in
Appendices \ref{A1} and \ref{A2}.

@@

This model is an extension of the cylindrical model introduced by
@abe2017tractable to the regression context. The density of the model for the
pair of the circular variable $\theta \in [0, 2\pi)$ and linear variable $y$ on
the positive real half-line $[0, + \infty)$ is:

@@

In a regression setup the outcome vector $(\theta_i, y_i)^t$ for each individual
$i = 1, \dots, n$, where $n$ is the sample size, is generated independently from
Equation (\ref{WeiSSVMdensity}). The parameter $\alpha > 0$ is a linear shape 
parameter, $\beta_i = \exp(\boldsymbol{x}_i^t\boldsymbol{\nu}) > 0$ is a linear 
scale parameter, $\mu_i = \eta_0 + 
2\tan^{-1}(\boldsymbol{z}_i^t\boldsymbol{\eta}) \in [0, 2\pi)$ is a circular 
location parameter, $\kappa > 0$ is a cicular concentration parameter and 
$\lambda \in [-1, 1]$ is a circular skewness parameter. The parameter 
$\boldsymbol{\nu}$ is a vector of $q$ regression coefficients and intercept
$\nu_j \in (-\infty, +\infty)$ where $j = 0, \dots, q$, for the prediction of
$y$. The parameter $\eta_0 \in [0, 2\pi)$ is the intercept and
$\boldsymbol{\eta}$, is a vector of $p$ regression coefficients $\eta_j \in 
(-\infty, +\infty)$, where $j = 1, \dots, p$, for the prediction of $\theta$. 
The vector $\boldsymbol{x}_i$ is a vector of predictor values for the prediction
of $y$ and $\boldsymbol{z}_i$ is a vector of predictor values for the prediction
of $\theta$.

@@

This model is an extension of the cylindrical model introduced by
@mastrantonio2018joint to the regression context. The model contains $p$
circular outcomes and $q$ linear outcomes. The circular outcomes 
$\boldsymbol{\Theta} = (\boldsymbol{\Theta}_1, \dots, \boldsymbol{\Theta}_p)$ 
are modeled together by a multivariate GPN distribution and the linear outcomes 
$\boldsymbol{Y} = (\boldsymbol{Y}_1, \dots, \boldsymbol{Y}_q)$  are modeled 
together by a multivariate skew normal distribution [@sahu2003new]. Because the
GPN distribution is modelled using a so-called augmented representation (as in
(\ref{projection}) and (\ref{gpnreg})) it is convenient to use a similar tactic
for modelling the multivariate skew normal distribution. As in
@mastrantonio2018joint we use the following representation of the linear
outcomes:

$$\boldsymbol{Y} = \boldsymbol{\mu}_y + \boldsymbol{\Lambda}\boldsymbol{D} + \boldsymbol{H},$$
where where $\boldsymbol{\mu}_y$ is a mean vector for the linear outcome 
$\boldsymbol{Y}$, $\boldsymbol{\Lambda} = \text{diag}(\boldsymbol{\lambda})$ is a $q 
\times q$ diagonal matrix with diagonal elements $\boldsymbol{\lambda} = 
\lambda_1, \dots, \lambda_q$ (the parameters that indicate the skewness of the 
linear outcomes), $\boldsymbol{D} \sim HN_q(\boldsymbol{0}_q, 
\boldsymbol{I}_q)$, a q-dimensional half normal distribution
[@olmos2012extension] and $\boldsymbol{H} \sim N_q(\boldsymbol{0}_q,
\boldsymbol{\Sigma}_y)$. This means that conditional on the auxiliary data
$\boldsymbol{D}$, $\boldsymbol{Y}$ is normally distributed with mean
$\boldsymbol{\mu}_y + \boldsymbol{\Lambda}\boldsymbol{D}$ and covariance matrix
$\boldsymbol{\Sigma}_y$. The joint density for $(\boldsymbol{Y},
\boldsymbol{D})^t$ is defined as follows:

@@

In our regression extension we have $i = 1, \dots, n$ observations of $p$ 
circular outcomes, $q$ linear outcomes and $k$ covariates. The mean in the 
density in (\ref{YDThetarjoint}) then becomes $\boldsymbol{\mu}_i = 
\boldsymbol{x}_i\boldsymbol{B}$ where $\boldsymbol{B}$ is a $k \times (2 
\times p + q)$ matrix with regression coeffients. We estimate the model using 
MCMC methods. A detailed description of these methods is given in Appendix
\ref{A3}.

@@

where $B$ is the amount of posterior samples and $\boldsymbol{\vartheta}^{(j)}$
are the posterior estimates of the model parameters for the $j^{th}$ iteration.
Note that although we fit the CL-PN, CL-GPN and joint GPN-SSN models using
Bayesian statistics, we do not take prior information (only the likelihood) into
account when assessing model fit. According to @BDA this is not necessary since
we are assessing the fit of a model to data only \textcolor{red}{AFMAKEN!}. For 
each of the cylindrical models we can then compute a PLSL for the circular and 
linear outcome by using the conditional log-likelihoods of the respective 
outcome.

@@

In this section we will analyze the teacher dataset with the help of
the four cylindrical models introduced in Section \ref{Models}. We will however
first give a more detailed description of the dataset.

@@

source("Posterior Sampling informative Prior CL-PN.R")

@@

source("Posterior Sampling CL-GPN.R")

@@

source("Posterior Sampling CL-GPN_multivariate.R")

@@

res_CLGPNM <- CLGPNM(theta.test, y.test, X.MGPN.test, its, p = 1, q = 1, 
                     theta.hold, y.hold, X.MGPN.hold)

@@

source("WeisSVM optimization.R")

@@

param <- c(1,1,1,1,1,1,0)*0.5

@@

The teacher data was collected between 2010 and 2015 and contains several repeated
measures on the IPC of 161 teachers gathered for the studies of
@vanderWant2015role, @Claessens2016side and @pennings2017interpersonal. 
The measurements were obtained using the QTI and taken in different years and in
different classes. For this paper we only consider one measurement, namely the
one for the first measurement occasion (2010) and the one for the largest class
if data for multiple classes were available. In addition to a variable `IPC`
containing the circular outcome and the 'length' of the IPC, the linear outcome,
a teachers' self-efficacy (\verb|SE|) will be used as covariate in the analysis.
After the removal of missing cases we end up with a sample of 148 teachers Table
\ref{Tableteacherdescriptives} shows descriptives for the dataset.

@@

#summary statistics
summary(Dat$Efficacy_SE)
summary(sqrt(Dat$lAgency^2 + Dat$lcommunion^2))
summary(as.circular(atan2(Dat$lAgency, Dat$lcommunion)%%(2*pi)))

sd(Dat$Efficacy_SE)
sd(sqrt(Dat$lAgency^2 + Dat$lcommunion^2))

@@

The variable \verb|SE| is thus used as a predictor for the linear and circular 
outcomes in all four cylindrical models. We will use the loglikelihoods of the
following conditional densities \textcolor{red}{CHRISTOPHE, MAG HIER WEL $\sim$ GEBRUIKT WORDEN?}for the computation of the PLSL criterion to
evaluate the model fit:

@@

source("Sample Abe-Ley.R")

>>>>

---
title: "A Comparison of Regression models for Cylindrical data in Psychology."
author: |
  | Jolien Cremers^[Corresponding author: j.cremers@uu.nl] $^1$, Helena J.M. Pennings$^{2,3}$,  Christophe Ley $^{4}$
  | $^1$Department of Methodology and Statistics, Utrecht University
  | $^2$TNO
  | $^3$Department of Education, Utrecht University
  | $^4$Department of Applied Mathematics, Computer Science and Statistics, Ghent University

bibliography: CircularData.bib
csl: apa.csl
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
  html_document: default
  word_document: default
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{multirow}
  - \usepackage{appendix}
  - \usepackage{color}
  - \usepackage{hyperref}
  - \usepackage{subcaption}

keywords: bla vla
documentclass: article
pandoc_args: --natbib
abstract: |
  Cylindrical dats is multivariate data which consists of a directional, in this paper circular, component   and a linear component. Examples of cylindrical data in psychology include human navigation (direction     and distance of movement), eye-tracking research (direction and length of saccades) and data from an       interpersonal circumplex (type and strength of interpersonal behavior). In this paper we adapt four        models for cylindrical data to include a regression of the circular and linear component onto a set of     covariates. Subsequently, we illustrate how to fit these models and interpret their results on a dataset   on the interpersonal behavior of teachers.
---
```{r inlude = FALSE, echo = FALSE}

@@

```{r, echo = FALSE}
library(circular)
library(bpnreg)
library(haven)
library(tikzDevice)
library(plotrix)
library(tidyr)
library(dplyr) #sample_n()
library(MASS) #mvrnorm()
library(kableExtra) #latex tables
library(plotrix)
```

@@

Cylindrical data are data that consist of a linear variable and a directional 
variable. In this paper, the directional variable is circular meaning that it 
consists of a single angle instead of a set of angles. A circular variable is 
different from a linear variable in the sense that it is measured on a different
scale. Figure \ref{circline} shows the difference between a circular scale 
(right) and a linear scale (left). The most important difference is that on a 
circular scale the datapoints 0$^\circ$ and 360$^\circ$ are connected and in 
fact represent the same number while on a linear scale the two ends, $-\infty$ 
and $\infty$ are not connected. This difference requires us to use different 
statistical methods for circular variables (see e.g. @fisher1995statistical for 
an introduction to circular data and @mardia2000directional,
@jammalamadaka2001topics and @ley2017modern for a more elaborate overview).

Cylindrical data occur in several fields of research, such as for instance in 
meteorology [@garcia2013exploring], ecology [@garcia2014test] or marine research
[@lagona2015hidden]. Several types of data in psychology are also of a 
cylindrical nature. For example, in research on human navigation in the field of
cognitive psychology both distance, a linear variable, and direction, a circular
variable, of movement are of interest [@chrastil2017rotational]. In eye-tracking
research, which can be used for investigating various cognitive processes (e.g. 
those involved in reading a text), we can also speak of cylindrical data. When 
measuring eye-movements we speak of the eye-movements themselves, the saccades, 
and fixations, the periods of time between movements when the eyes are looking 
at one point. Of the saccades both the direction, a circular variable, and the 
duration, a linear variable, are of interest (for a review of eye-tracking 
research see @rayner200935th). Data from circumplex measurement instruments,
e.g. the interpersonal circumplex as used in personality psychology, are also of
a cylindrical nature (see Section \ref{Example} for a more detailed
explanation).

In this paper we will discuss how a correct statistical treatment of such 
cylindrical data can lead to new insights. In particular we will show how 
cylindrical models pave the way for circular-linear and linear-circular 
regression. We will do this for a motivating example, the teacher data,  from
the field of educational psychology . In this example, apart from modelling the
relation between the linear and circular component of a cylindrical variable we
would also like to predict the two components from a set of covariates in a
regression model. The teacher data will be further introduced in Sections
\ref{Example} and \ref{DataAnalysis}.

\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{Plots/circline.pdf}
\caption{The difference between a linear scale (left) and a circular scale (right).}
\label{circline}
\end{figure}

As is the case for circular data, the analysis of cylindrical data requires 
special methods. Several methods have been put forward to model the relation 
between the linear and circular component of a cylindrical variable. Some of 
these are based on regressing the linear component onto the circular component 
using the following type of relation: $$y = \beta_0 + \beta_1*\cos(\theta) + 
\beta_2*\sin(\theta)+ \epsilon,$$ where $y$ is the linear component and $\theta$
the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a 
different way, e.g. by specifying a multivariate model for several linear and 
circular variables and modelling their covariance matrix 
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. For 
example, @abe2017tractable introduce a cylindrical distribution based on a 
Weibull distribution for the linear component and a sine-skewed von Mises 
distribution for the circular component and link these through their shape and 
concentration parameters respectively. However, none of the methods that have 
been proposed thus far include additional covariates onto which both the
circular and linear component are regressed.

Our aim in this paper is to fill this gap in the literature by adapting four
existing cylindrical methods in such a way that they include a regression of
both the linear and circular component of a cylindrical variable onto a set of
covariates. First however, we will introduce the teacher data in Section
\ref{Example}. In Section \ref{Models} we introduce the four modified models for
cylindrical data that we use to analyze the data from the motivating example. We
also choose a model selection criterion to compare the models. The teacher data
will be analysed in Section \ref{DataAnalysis}. The paper will be concluded with
a discussion in Section \ref{Discussion}. The Appendix contains technical
details.

@@

The motivating example for this article comes from the field of educational 
psychology and was collected for the studies on classroom climate of
@pennings2017interpersonal, @Claessens2016side and @vanderWant2015role. An 
indicator of classroom climate is the students' perceptions of their teachers'
interpersonal behavior. These interpersonal perceptions, both in educational
psychology as well as in other areas of psychology, can be measured using
circumplex measurement instruments (see
@horowitz2010handbook for an overview of many such instruments).

The Questionnaire on Teacher Interaction (QTI) [@wubbels2006interpersonal] is
one such circumplex measurement instrument. It is used to study student
perceptions of their teachers' interpersonal behavior and contains
items that load on two dimensions: Agency and Communion. Agency refers to the
degree of power or control a teacher exerts in interaction with his/her
students. Communion refers to the degree of friendliness or affiliation a
teacher conveys in interaction with his/her students. The loadings on the two 
dimensions of the QTI can be placed in a two-dimensional space formed by Agency
(vertical) and Communion (horizontal), see Figure \ref{QTI}. Different parts of
this space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. We call the two-dimensional space the interpersonal circle (IPC).
The idea is that the IPC is ``a continuous order with no beginning or end''
[@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering and
the IPC is therefore often called the interpersonal circumplex. The ordering
also implies that scores on the IPC could be viewed as a circular variable.

\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{Plots/IPC-T.png}
\caption{The interpersonal circle for teachers (IPC-T). The words presented in
the circumference of the circle are anchor words to describe the type of
behavior located in each part of the IPC.}
\label{QTI}
\end{figure}

@Cremers2018Assessing show how data from the IPC can be considered circular data
and analyzed as such using a regression model. The two-dimension scores Agency
and Communion can be converted to a circular score using the two-argument
arctangent function in \eqref{PredVal}, where $A$ represents a score on
the Agency dimension and $C$ represents a score on the Communion dimension. The
resulting circular variable $\theta$ can then be modeled. However, when
two-dimensional data are converted to the circle we lose some information, the
length of the two-dimensional vector $(A, C)$, \emph{i.e.}, its euclidean norm
$\mid\mid (A, C) \mid\mid$. This length represents the strength of the type of
interpersonal behavior a teacher shows towards his/her students. In a
cylindrical model we are able to incorporate this information, and model a
circular variable $\theta$ together with a linear variable corresponding to
$\mid\mid (A, C) \mid\mid$. This leads to an improved analysis of data from the
IPC. In the next section we introduce several models that can be used for more
accurate and informative regression analysis on the teacher data. Descriptives
for the teacher data are given in Section \ref{DataAnalysis}.

@@

#center and remove missings for self-efficacy
Dat <- Dat %>% drop_na(Efficacy_CM)
Dat$SEc <- Dat$Efficacy_CM- mean(Dat$Efficacy_CM)

@@

In this section we introduce four cylindrical regression models to contain 
predictors for the linear and circular outcomes, $\Theta$ and $Y$. We extend the models from @mastrantonio2018joint and
@abe2017tractable. Additionally, we introduce two models where the relation
between $\Theta$ and $Y$ is modelled as follows (following
@mastrantonio2015bayesian):

@@

In both of these models the relation between $\Theta \in [0, 2\pi)$ and $Y\in 
(-\infty, + \infty)$ is specified as in \eqref{circlinlink} with the following
conditional distribution:

@@

In a regression setup the outcome $\theta_i$ for each individual $i = 1, \dots,
n$, where $n$ is the sample size, is generated independently from \eqref{pnreg}.
The mean vector $\boldsymbol{\mu}_i \in \mathbb{R}^2$ is defined as
$\boldsymbol{\mu}_i = \boldsymbol{z}_i\boldsymbol{B}$. The vector 
$\boldsymbol{z}_i$ is a a vector of $p$ covariate values and $\boldsymbol{B} = 
(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})$ contain the regression 
coefficients. Note however that the dimensions of $\boldsymbol{\beta}^{I}$ and 
$\boldsymbol{\beta }^{I}$ need not necessarily be the same and we are thus 
allowed to have a different set of predictor variables and vectors 
$\boldsymbol{z}_i^I$ and $\boldsymbol{z}_i^{II}$ for the two components of 
$\boldsymbol{\mu}_i$.

@@

where $\boldsymbol{\Sigma} = \begin{bmatrix} \tau^2 + \rho^2  & \rho\\ \rho & 1 
\end{bmatrix}$, $\boldsymbol{u}= \begin{bmatrix} \cos \theta \\ \sin \theta 
\end{bmatrix}$. In a regression setup the outcome $\theta_i$ for each individual
is generated independently from \eqref{gpnreg}. The mean vector 
$\boldsymbol{\mu}_i \in \mathbb{R}^2$ is defined as $\boldsymbol{\mu}_i = 
\boldsymbol{z}_i(\boldsymbol{\beta}^{I}, \boldsymbol{\beta}^{II})$. The vector 
$\boldsymbol{z}_i$ is a a vector of $p$ covariate values and each 
$\boldsymbol{\beta}^{k}$ is a vector with two regression coefficients, one for 
each of the two components of $\boldsymbol{\mu}_i$. Note that for the CL-GPN 
model we do need to have the same predictors for both components of 
$\boldsymbol{\mu}_i$.

Both cylindrical models introduced here are estimated using MCMC methods based
on @nunez2011bayesian, @wang2012directional and @hernandez2016general for the 
regression of the circular outcome. A detailed description of the Bayesian
estimation and MCMC samplers can be found in Appendices \ref{A1} and \ref{A2}.

@@

This model is an extension of the cylindrical model introduced by
@abe2017tractable to the regression context. The joint density of $\Theta$ and
$Y$, in this model defined on the positive real half-line $[0, + \infty)$, is:

@@

In a regression setup the outcome vector $(\theta_i, y_i)^t$ for each individual
is generated independently from \eqref{WeiSSVMdensity}. The parameters $\alpha >
0$ and $\beta_i = \exp(\boldsymbol{x}_i^t\boldsymbol{\nu}) > 0$ are linear shape
and scale parameters, $\mu_i = \eta_0 + 
2\tan^{-1}(\boldsymbol{z}_i^t\boldsymbol{\eta}) \in [0, 2\pi)$, $\kappa > 0$ and
$\lambda \in [-1, 1]$ are circular location, concentration and skewness 
parameters. The parameter $\boldsymbol{\nu}$ is a vector of $q$ regression 
coefficients $\nu_j \in (-\infty, +\infty)$ for the prediction of $y$ where $j =
0, \dots, q$ and $\nu_0$ is the intercept. The parameter $\eta_0 \in [0, 2\pi)$
is the intercept and $\boldsymbol{\eta}$ is a vector of $p$ regression
coefficients $\eta_j \in (-\infty, +\infty)$ for the prediction of $\theta$
where $j = 1, \dots, p$. The vector $\boldsymbol{x}_i$ is a vector of predictor
values for the prediction of $y$ and $\boldsymbol{z}_i$ is a vector of predictor
values for the prediction of $\theta$.

@@

This model is an extension of the cylindrical model introduced by
@mastrantonio2018joint to the regression context. It contains $p$ circular
outcomes and $q$ linear outcomes. The circular outcomes $\boldsymbol{\Theta} = 
(\boldsymbol{\Theta}_1, \dots, \boldsymbol{\Theta}_p)$ are modeled together by a
multivariate GPN distribution and the linear outcomes $\boldsymbol{Y} = 
(\boldsymbol{Y}_1, \dots, \boldsymbol{Y}_q)$  are modeled together by a 
multivariate skew normal distribution [@sahu2003new]. Because the GPN 
distribution is modelled using a so-called augmented representation (as in 
\eqref{projection} and \eqref{gpnreg}) it is convenient to use a similar tactic
for modelling the multivariate skew normal distribution. Following
@mastrantonio2018joint the linear outcomes are represented as:

$$\boldsymbol{Y} = \boldsymbol{\mu}_y + \boldsymbol{\Lambda}\boldsymbol{D} + \boldsymbol{H},$$
where where $\boldsymbol{\mu}_y$ is a mean vector for the linear outcome 
$\boldsymbol{Y}$, $\boldsymbol{\Lambda} = \text{diag}(\boldsymbol{\lambda})$ is a $q 
\times q$ diagonal matrix with diagonal elements $\boldsymbol{\lambda} = 
\lambda_1, \dots, \lambda_q$ (skewness parameters), $\boldsymbol{D} \sim HN_q(\boldsymbol{0}_q, 
\boldsymbol{I}_q)$, a q-dimensional half normal distribution
[@olmos2012extension], and $\boldsymbol{H} \sim N_q(\boldsymbol{0}_q,
\boldsymbol{\Sigma}_y)$. This means that conditional on the auxiliary data
$\boldsymbol{D}$, $\boldsymbol{Y}$ is normally distributed with mean
$\boldsymbol{\mu}_y + \boldsymbol{\Lambda}\boldsymbol{D}$ and covariance matrix
$\boldsymbol{\Sigma}_y$. The joint density for $(\boldsymbol{Y},
\boldsymbol{D})^t$ is defined as follows:

@@

In our regression extension we have $i = 1, \dots, n$ observations of $p$ 
circular outcomes, $q$ linear outcomes and $k$ covariates. The mean in the 
density in \eqref{YDThetarjoint} then becomes $\boldsymbol{\mu}_i = 
\boldsymbol{x}_i\boldsymbol{B}$ where $\boldsymbol{B}$ is a $k \times (2 
\times p + q)$ matrix with regression coeffients. We estimate the model using 
MCMC methods. A detailed description of these methods is given in Appendix
\ref{A3}.

@@

where $B$ is the amount of posterior samples and $\boldsymbol{\vartheta}^{(j)}$ 
are the posterior estimates of the model parameters for the $j^{th}$ iteration. 
Note that although we fit the CL-PN, CL-GPN and joint GPN-SSN models using 
Bayesian statistics, we do not take prior information into account when 
assessing model fit. According to @BDA \textcolor{red}{(p. ???)} this is not 
necessary since we are assessing the fit of a model to data, the holdout set,
only. They argue that the prior in such case is only of interest for estimating
the parameters of the model but not for determining the predictive accuracy. 
For each of the cylindrical models we can then compute a PLSL for the circular
and linear outcome by using the conditional log-likelihoods of the respective
outcome.

@@

In this section we analyze the teacher data with the help of the four
cylindrical models from Section \ref{Models}. First however, we give a more
detailed description of the dataset.

@@

source("R-code/Posterior Sampling CL-PN.R")

@@

source("R-code/Posterior Sampling CL-GPN.R")

@@

source("R-code/Posterior Sampling Joint GPN-SSN.R")

@@

res_CLGPNM <- JGPNSSN(theta.test, y.test, X.MGPN.test, its, p = 1, q = 1, 
                      theta.hold, y.hold, X.MGPN.hold)

@@

source("R-code/Abe-Ley optimization.R")

@@

param <- c(1,1,1,1,1,1,0)*0.9

@@

The teacher data was collected between 2010 and 2015 and contains several 
repeated measures on the IPC of 161 teachers. Measurements were obtained using 
the QTI and taken in different years and classes. For this paper we only 
consider one measurement, the first occasion (2010) and largest class if data 
for multiple classes were available. In addition to the score on the IPC, the 
circular outcome, and the strength of the score on the IPC, the linear outcome, 
a teachers' self-efficacy (\verb|SE|) concerning classroom management ia used as
covariate in the analysis. After the removal of missings we have a sample of 148
teachers of which 10\% (n = 15) is used as a holdout set for the computation of
the PLSL. Table \ref{Tableteacherdescriptives} shows descriptives
for the dataset. Figure \ref{dataplot} is a scatterplot showing the relation
between the linear and circular outcome of the teacher data.

```{r dataplot, cache = TRUE, dev = "tikz",fig.ext="svg", echo = FALSE, results = FALSE}

tikz("Plots/dataplot.tex", standAlone =TRUE, height = 3.5, width = 6, pointsize = 12, engine = "pdftex")

plot(y.test, theta.test*(180/pi), 
     ylab = "IPC location", xlab = "IPC strength",
     bty = "n", ylim = c(-180, 180))

dev.off()
tools::texi2dvi("Plots/dataplot.tex", pdf=TRUE)

@@

#summary statistics
summary(Dat$Efficacy_CM)
summary(sqrt(Dat$lAgency^2 + Dat$lcommunion^2))
summary(as.circular(atan2(Dat$lAgency, Dat$lcommunion)%%(2*pi)))

sd(Dat$Efficacy_CM)
sd(sqrt(Dat$lAgency^2 + Dat$lcommunion^2))

@@

We use the loglikelihoods of the following conditional densities for the
computation of the PLSL:

@@

source("R-code/Sample Abe-Ley.R")
